---
title: "Overview of the 'fastglm' Package"
author: "Jared Huling"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Overview of the 'fastglm' Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  echo = TRUE,
  eval = TRUE
)
```

The *fastglm* package is intended as a **fast** *and* **stable** alternative to `stats::glm()` for fitting generalized linear models. It is compatible with `R`'s `family` objects (see `?family`) and produces fitted objects whose downstream methods --- `summary()`, `vcov()`, `predict()`, `coef()`, `residuals()`, `logLik()` --- behave exactly like those for a `glm`. This vignette walks through the main functionality of the package at a higher level than the other vignettes; it is intended as a single entry point for a new user.

```{r}
library(fastglm)
```

# Fitting a GLM

The main package function is `fastglm()`. It takes a numeric design matrix `x`, a response `y`, and an `R` `family` object:

```{r}
data(esoph)
x <- model.matrix(~ agegp + unclass(tobgp) + unclass(alcgp),
                  data = esoph)
y <- cbind(esoph$ncases, esoph$ncontrols)

fit <- fastglm(x, y, family = binomial(link = "cloglog"))
summary(fit)
```

`fastglm()` does not accept a formula directly, it operates on a pre-built design matrix. To use a formula and a data frame, pass `fastglm.fit` as the fitting function to base `glm()`:

```{r}
fit2 <- glm(cbind(ncases, ncontrols) ~ agegp + unclass(tobgp) + unclass(alcgp),
            data = esoph, family = binomial(link = "cloglog"),
            method = fastglm.fit)
```

# Decomposition methods

The IRLS algorithm reduces every iteration to a weighted least-squares problem. *fastglm* supports six different matrix decompositions for solving that WLS step, all from *RcppEigen* (Bates and Eddelbuettel, 2013); the choice trades off speed against numerical stability and rank-revealing behaviour:

| `method` | decomposition |
|---|---|
| `0` | column-pivoted Householder QR (default; rank-revealing) |
| `1` | unpivoted Householder QR |
| `2` | LLT Cholesky |
| `3` | LDLT Cholesky |
| `4` | full-pivoted Householder QR |
| `5` | bidiagonal divide-and-conquer SVD |

The default (`method = 0`) is the safe choice: it is rank-revealing, so it handles aliased / collinear columns gracefully. The Cholesky methods (`2` and `3`) are roughly 3–4× faster but assume full column rank.

```{r}
set.seed(123)
n <- 5000; p <- 30
x <- matrix(rnorm(n * p), n, p)
y <- rbinom(n, 1, plogis(x %*% rnorm(p) * 0.05))

system.time(f0 <- fastglm(x, y, family = binomial()))                 # default
system.time(f2 <- fastglm(x, y, family = binomial(), method = 2))     # LLT
system.time(f3 <- fastglm(x, y, family = binomial(), method = 3))     # LDLT
```

# Stability

*fastglm* does not compromise computational stability for speed. It uses a step-halving safeguard following Marschner (2011) and starts from better-initialised values than `glm()` or `glm2::glm2()`, so it tends to converge in cases where the standard IRLS algorithm fails. See `vignette("fastglm", package = "fastglm")` for a Gamma-with-`sqrt`-link example where `glm()` does not converge but `fastglm()` does.

# Inference: `vcov()`, robust SE, predictions

The fitted object stores the unscaled covariance directly (no refitting), and the standard `vcov()` / `summary()` machinery works as expected:

```{r}
fit <- fastglm(x, y, family = binomial(), method = 2)

V    <- vcov(fit)
dim(V)

## standard errors agree with the diagonal of vcov()
all.equal(sqrt(diag(V)), unname(coef(summary(fit))[, "Std. Error"]))
```

Heteroskedasticity-consistent and cluster-robust covariance matrices are
available via `sandwich::vcovHC()` and `sandwich::vcovCL()` — `fastglm`
registers methods on those generics, so loading `sandwich` is all that is
required:

```{r}
library(sandwich)

V_hc <- vcovHC(fit, type = "HC0")
V_hc[1:3, 1:3]

cluster <- rep(seq_len(20), length.out = n)
V_cl    <- vcovCL(fit, cluster = cluster, type = "HC1")
V_cl[1:3, 1:3]
```

Results are numerically identical to `sandwich::vcovHC()` and
`sandwich::vcovCL()` applied to a `glm` fit (Zeileis, Köll, and Graham, 2020)
to floating-point precision. They work for sparse and `big.matrix` fits as
well, since the full design matrix is preserved on the fitted object (as a
reference, not copied).

`predict()` supports `se.fit = TRUE`:

```{r}
xnew <- x[1:5, , drop = FALSE]
predict(fit, newdata = xnew, type = "response", se.fit = TRUE)
```

# Sparse, big.matrix, and streaming designs

For designs that are sparse, that live on disc, or that have to be built from a parquet / *arrow* / *DuckDB* source, *fastglm* provides three large-data paths that share a common streaming kernel and produce identical results:

- `Matrix::dgCMatrix`: pass directly to `fastglm()`. Useful for one-hot encoded categoricals and high-dimensional sparse designs.
- `bigmemory::big.matrix`: pass directly to `fastglm()`. The matrix is read in row-blocks and never fully materialised in memory.
- `fastglm_streaming(chunk_callback, n_chunks, family)`: a user-supplied closure yields one row-block per call. The right path for fitting on a parquet dataset, *DuckDB* query, or any external columnar store.

See `vignette("large-data-fastglm", package = "fastglm")` for a detailed walk through all three paths. A short example:

```{r}
n <- 4000
X <- cbind(1, matrix(rnorm(n * 3), n, 3))
y <- rbinom(n, 1, plogis(X %*% c(0.2, 0.4, -0.2, 0.3)))

chunk_size <- 1000
chunks <- function(k) {
    idx <- ((k - 1) * chunk_size + 1):(k * chunk_size)
    list(X = X[idx, , drop = FALSE], y = y[idx])
}

fit_stream <- fastglm_streaming(chunks, n_chunks = 4, family = binomial())
fit_full   <- fastglm(X, y, family = binomial(), method = 2)

max(abs(coef(fit_stream) - coef(fit_full)))
```

# Native families

For the most commonly-used `family`/`link` combinations, *fastglm* dispatches `variance()`, `mu.eta()`, `linkinv()`, and `dev.resids()` to inline C++ implementations rather than calling back into `R` once per IRLS iteration. The covered combinations are:

- gaussian (identity, log, inverse)
- binomial (logit, probit, cloglog, log)
- poisson (log, identity, sqrt)
- Gamma (log, inverse, identity)
- inverse.gaussian (1/mu^2, log, identity, inverse)

Detection is automatic, if the `family` object matches one of the above, the native path is used; otherwise *fastglm* falls back to the `R`-callback path used in earlier versions. The native path is meaningfully faster on large `n` because it eliminates the per-iteration round-trip to `R` for each of the four family functions:

```{r, eval = TRUE}
library(microbenchmark)

set.seed(7)
n <- 50000; p <- 30
x <- matrix(rnorm(n * p), n, p)
y <- rpois(n, exp(x %*% rnorm(p) * 0.05 + 1))

## force the R-callback path by giving the family object an unrecognised name
disguise <- function(fam) { fam$family <- paste0(fam$family, "*"); fam }

mb <- microbenchmark(
    native   = fastglm(x, y, family = poisson(),            method = 2),
    callback = fastglm(x, y, family = disguise(poisson()),  method = 2),
    times = 5L
)
print(mb)
```

The *fastglm* package switches transparently between the two paths; user code does not change.

# Negative binomial, hurdle, zero-inflation, and Firth logistic

A handful of regressions that come up often in practice are not standard GLMs but build on the same IRLS machinery. *fastglm* provides dedicated entry points for these:

- **`fastglm_nb()`** — negative-binomial regression with the dispersion `theta` estimated jointly with `beta`. Plays the same role as `MASS::glm.nb()` but runs the joint `(beta, theta)` MLE entirely in C++ (IRLS for `beta`, Brent for `theta`).
- **`firth = TRUE`** — a flag on `fastglm()` / `fastglm.fit()` that activates Firth's bias-reducing penalty on the score. Useful for binary logistic regression under separation or in small samples; analogous to `logistf::logistf()`.
- **`fastglm_hurdle()`** — a two-part count model: a binary regression for whether `y > 0`, plus a zero-truncated Poisson or NB regression on the positive subset. The two parts factorize, and both are fit by the same C++ IRLS solver. Same model as `pscl::hurdle()`.
- **`fastglm_zi()`** — a zero-inflated Poisson or NB regression: a binary inflation component overlaid on the original count distribution, fit by an EM algorithm in C++ with closed-form posterior responsibilities and an analytical observed-information `vcov`. Same model as `pscl::zeroinfl()`.

All four take the standard *fastglm* tuning options (`method`, `tol`, `maxit`) and return objects with the usual `coef()`, `vcov()`, `predict()`, `logLik()` methods. `fastglm_hurdle()` and `fastglm_zi()` use the `Formula` package's two-RHS syntax `y ~ x1 + x2 | z1 + z2` to specify different designs for the count and zero parts. See `vignette("count-firth-fastglm", package = "fastglm")` for examples and timing comparisons against the canonical reference packages.

# Three R entry points

There are three `R`-side functions to fit a GLM with *fastglm*. They all call into the same C++ solver, but differ in how much of base R's machinery they wrap around it:

- **`fastglm()`** is the top-level S3 function. It returns an object of class `"fastglm"` with deviance, AIC, null deviance, residuals, etc. populated --- the right entry point for interactive use.
- **`fastglm.fit()`** is designed to be passed as `method = fastglm.fit` to base `glm()`. The returned object inherits from `"fastglmFit"` so that `glm()`'s formula/data handling, `update()`, `predict()`, etc. all work on top.
- **`fastglmPure()`** is the lowest-level wrapper: no dispersion, no AIC, no null-deviance computation. Use this when calling *fastglm* from another package and you only need the coefficients.

# References

- Marschner, I. C. (2011). glm2: Fitting generalized linear models with convergence problems. *The R Journal*, 3(2), 12–15. <doi:10.32614/RJ-2011-012>

- Bates, D. and Eddelbuettel, D. (2013). Fast and elegant numerical linear algebra using the RcppEigen package. *Journal of Statistical Software*, 52(5), 1–24. <doi:10.18637/jss.v052.i05>

- Zeileis, A., Köll, S., and Graham, N. (2020). Various versatile variances: An object-oriented implementation of clustered covariances in R. *Journal of Statistical Software*, 95(1), 1–36. <doi:10.18637/jss.v095.i01>
